Yeast RNA-seq 48 sample simulation study - II

### 2018-08-15 datasets/RNA-Seq/yeast-simulation-bimodalalternative.Rmd
Sample observed test statistics using bimodal sampling weights

In this Rmd we will carry out simulations for multiple
replicates, and plot results averaging over the replications. In contrast to
`yeast-simulation.Rmd` we will use an alternate simulation framework for generating
the effect sizes of the DE genes. Here we will use the fold changes of only genes
that were DE in a comparison of wild type versus knockout (with sample size 10).
Since this results in a distribution of effect sizes (log2 fold changes) under the
alternative that does not have a mode at zero, the assumptions for ashq are violated.
For this reason, we do not include this method in the benchmark comparisons.

@@

dds <- DESeq(subset)
resultsNames(dds) # lists the coefficients
resFULL <- results(dds, name="condition_WT_vs_Snf2", independentFiltering = F)

head(resFULL)

sum(resFULL$padj < 0.05, na.rm=T)
```

@@

```{r}
data.frame(resFULL[resFULL$padj < 0.05,]) %>% 
  ggplot(aes(x = stat)) +
  geom_density() +
  xlab("Test Statistic")

data.frame(resFULL[resFULL$padj < 0.05,]) %>% 
  ggplot(aes(x = log2FoldChange)) +
  geom_density() +
  xlab("Effect Size (log2 Fold Change") 
```

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resFULL), 
                             pval=resFULL$pvalue, 
                             SE=resFULL$lfcSE,
                             ind_covariate = resFULL$baseMean, 
                             effect_size=resFULL$log2FoldChange, 
                             test_statistic=resFULL$stat,
                             pzero=rowSums(counts(subset)==0)/ncol(counts(subset))))

@@

Subset on only significant DE genes (FDR < 0.05) to draw fold changes from.

```{r}
resFULL <- resFULL[resFULL$padj < 0.05,]
```

@@

Build common bench design. We also add in Scott's FDR Regression (both
`nulltype = "empirical"` and `nulltype = "theoretical"`)
since our test statistics are approximately t-distributed. We also 
drop the ashq method since its assumptions are violated here.

```{r}
bd <- initializeBenchDesign()
bd <- addBMethod(bd, "fdrreg-t",
                     FDRreg::FDRreg,
                     function(x) { x$FDR },
                     z = test_statistic,
                     features = model.matrix( ~  splines::bs(ind_covariate, df = 3) - 1),
                     nulltype = 'theoretical',
                     control = list(lambda = 0.01))
bd <- addBMethod(bd, "fdrreg-e",
                     FDRreg::FDRreg,
                     function(x) { x$FDR },
                     z = test_statistic,
                     features = model.matrix( ~  splines::bs(ind_covariate, df = 3) - 1),
                     nulltype = 'empirical',
                     control = list(lambda = 0.01))
bd <- dropBMethod(bd, "ashq")
```

@@

@@

    # randomly sample a log2FC from original FCs (without regard to DE)
    counts_new <- counts(dds_test)
    log2FC <- rep(0, nrow(dds_test))
    log2FC[DE] <- resFULL$log2FoldChange[sample(1:nrow(resFULL), nDE, 
                                                replace = TRUE)]
    log2FC[DE] <- pmax(1, abs(rnorm(nDE, 2)))

@@

    counts_new[down,colData(dds_test)$fake==unique(refcond[down])] <- 
        counts(dds_test)[down, colData(dds_test)$fake==unique(refcond[down])] *
        2^log2FC
    counts_new[-down,colData(dds_test)$fake==unique(refcond[-down])] <- 
        counts(dds_test)[-down, colData(dds_test)$fake==unique(refcond[-down])] *
        2^log2FC
    counts_new <- apply(counts_new, 2, as.integer)

>>>>

In this Rmd we will carry out simulations for multiple
replicates, and plot results averaging over the replications. In contrast to
`yeast-simulation.Rmd` we will use an alternate simulation framework for generating
the effect sizes of the DE genes. Here we will use the fold changes of only genes
that were DE in a comparison of wild type versus knockout (with sample size 10).
Since this results in a distribution of effect sizes (log2 fold changes) under the
alternative that does not have a mode at zero, the assumptions for ashq are violated.
However, since this assumption is impossible to check, we still include this 
method in the benchmark comparisons.

@@

dds <- DESeq(subset)
resultsNames(dds) # lists the coefficients
res_10 <- results(dds, name="condition_WT_vs_Snf2", independentFiltering = F)

head(res_10)

sum(res_10$padj < 0.05, na.rm=T)
```

@@

```{r}
data.frame(res_10[res_10$padj < 0.5]) %>% 
  ggplot(aes(x = stat)) +
  geom_density() +
  xlab("Test Statistic")

data.frame(res_10[res_10$padj < 0.5]) %>% 
  ggplot(aes(x = log2FoldChange)) +
  geom_density() +
  xlab("Effect Size (log2 Fold Change)") 
```

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(res_10), 
                             pval=res_10$pvalue, 
                             SE=res_10$lfcSE,
                             ind_covariate = res_10$baseMean, 
                             effect_size=res_10$log2FoldChange, 
                             test_statistic=res_10$stat,
                             pzero=rowSums(counts(subset)==0)/ncol(counts(subset))))

@@

@@

Build common bench design. We also add in Scott's FDR Regression 
since our test statistics are approximately t-distributed. 
Note that the assumption of FDRreg-empirical that the distribution of non-null test 
statistics does not have a significant mass at zero is violated here, but
we still include fdrreg-e in the evaluation since this assumption is impossible to check
outside of a simulation setting (since it is not possible to know which tests are 
non-null). Thus, we'd like to evaluate how results change 
when the assumption is violated, since it's plausible that it is violated in the 
case studies.

```{r}
bd <- initializeBenchDesign()
bd <- addBMethod(bd, "fdrreg-t",
                     FDRreg::FDRreg,
                     function(x) { x$FDR },
                     z = test_statistic,
                     features = model.matrix( ~  splines::bs(ind_covariate, df = 3) - 1),
                     nulltype = 'theoretical',
                     control = list(lambda = 0.01))
bd <- addBMethod(bd, "fdrreg-e",
                     FDRreg::FDRreg,
                     function(x) { x$FDR },
                     z = test_statistic,
                     features = model.matrix( ~  splines::bs(ind_covariate, df = 3) - 1),
                     nulltype = 'empirical',
                     control = list(lambda = 0.01))
```

# Empirical analysis of 5 vs 5 (non-null)

Here we'll carry out an analysis of a subset of the full dataset, comparing the controls to the
knockout samples. The fold changes observed in this comparison will be used when 
generating the non-null simulated data in the following sections. Here we use a subset
of the samples (5 samples in each group) to mimic the sample sizes we use in 
simulation.

First we run DESeq2, and subset results on genes with FDR < 0.05.

```{r}
# results on full set
set.seed(728)
subset.wt <- dds_full[,colData(dds_full)$condition == "WT"]
subset.sn <- dds_full[,colData(dds_full)$condition == "Snf2"]
subset <- cbind(subset.wt[, sample(1:ncol(subset.wt), 5)],
                subset.sn[, sample(1:ncol(subset.sn), 5)])

dds <- DESeq(subset)
resultsNames(dds) # lists the coefficients
res_5 <- results(dds, name="condition_WT_vs_Snf2", independentFiltering = F)

head(res_5)

sum(res_5$padj < 0.05, na.rm=T)
```

```{r}
data.frame(res_5[res_5$padj < 0.5]) %>% 
  ggplot(aes(x = stat)) +
  geom_density() +
  xlab("Test Statistic")

data.frame(res_5[res_5$padj < 0.5]) %>% 
  ggplot(aes(x = log2FoldChange)) +
  geom_density() +
  xlab("Effect Size (log2 Fold Change)") 
```

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(res_5), 
                             pval=res_5$pvalue, 
                             SE=res_5$lfcSE,
                             ind_covariate = res_5$baseMean, 
                             effect_size=res_5$log2FoldChange, 
                             test_statistic=res_5$stat,
                             pzero=rowSums(counts(subset)==0)/ncol(counts(subset))))

```{r, width=15, height=15}
strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy=23)
```

@@

First, here's a function that will generate sampling weights for test statistics
of DE genes that downweights observations near zero, as well as extremely large values.

```{r}
stat_wt <- function(z){
  pos <- pmax(0,dgamma(z,4.5, 1-1e-4))
  neg <- pmax(0,dgamma(-z,4.5, 1-1e-4))

  res <- pos
  res[z < 0] <- neg[z < 0]
  return(res)
}

# Plot weights by test statistic value
df = data.frame(z = seq(-10,10, by = 0.1))
df$w = stat_wt(df$z)
ggplot(data=df, aes(x=z, y=w)) +
  geom_line() +
  ylab("Sampling Weight") +
  xlab("Test statistic value")
```

@@

    # randomly sample a test statistic from original test statistics, with probability
    # weights that downweight observations near zero 
    counts_new <- counts(dds_test)
    stat <- rep(0, nrow(dds_test))

    # assume fixed standard error (median observed) and calculate FC from stat
    if(sampleSize == 5){
      stat[DE] <- sample(res_5$stat, nDE, prob = stat_wt(res_5$stat))
      log2FC <- stat * median(res_5$lfcSE)
    }else if (sampleSize == 10){
      stat[DE] <- sample(res_10$stat, nDE, prob = stat_wt(res_10$stat))
      log2FC <- stat * median(res_10$lfcSE)
    }else{
      stop("Only sample sizes 5 and 10 are currently supported with pre-",
           "computed fold changes for sampling from.")
    }

@@

    counts_new[down,colData(dds_test)$fake==unique(refcond[down])] <- 
        counts(dds_test)[down, colData(dds_test)$fake==unique(refcond[down])] *
        2^log2FC[down]
    counts_new[-down,colData(dds_test)$fake==unique(refcond[-down])] <- 
        counts(dds_test)[-down, colData(dds_test)$fake==unique(refcond[-down])] *
        2^log2FC[-down]
    counts_new <- apply(counts_new, 2, as.integer)
